Direction-aware functional class scoring enrichment analysis of Infinium DNA methylation data

### 2024-02-14 manuscript/manuscript.Rmd
img

![**Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.**
Upper panels show the number of significant pathways identified in the smaller sample that were consistent
with the full group for LAM (left) and GSA (right).
Middle panels show the number of significant pathways identified in the smaller sample that were inconsistent
with the full group.
Lower panels show the proportion of significant pathways identified that were inconsistent with the full
group.
Left panels correspond to LAM, and right panels to GSA methods.](img/fig4.svg)

@@

![**Figure 5. Integration of gene expression data with methylation pathways.**
(A,B) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from gene expression and GSA results (A)
and LAM (B) results.
(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
Top 20 gene sets shown with the largest S distance after FDR filtering at 0.05.](img/fig5.svg)

@@

![**Figure 6. Pathway-level DNA methylation alterations with chronological age.**
(A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
(B) Mitch pathway enrichment scores (S distances) in discovery and replication studies.
Pathways with MANOVA FDR<0.05 are shown in red while others are shaded grey.
(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
Red indicates increasing methylation and blue shows lower methylation.
(D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes
in both discovery and replication studies.](img/fig6.svg)

@@

![**Figure 7. Pathway-level DNA methylation differences in natural and IVF conceived infants.**
(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (HM450K) and Novakovic (EPIC) studies.
(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies.
Pathways with MANOVA FDR<0.05 are shown in red while others are shaded grey.
(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
Red indicates higher methylation and blue shows lower methylation.
(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of
member genes in both Estill and Novakovic studies.](img/fig7.svg)

@@

![**Figure 8. Differential pathway methylation associated with prevalence of 14 common disease states.**
(A) A bar plot showing the number of statistically significant pathways with higher and lower methylation identified in each prevalent condition (FDR<0.05).
(B) A heatmap of S scores for selected pathways across 14 common prevalent disease states.
Stars indicate that the pathway was identified as being among the top five differentially methylated pathways in each direction for each condition.
](img/fig8.svg)

>>>>

![**Figure 4. Sensitivity of LAM and GSA methods with real cancer array data.**
Upper panels show the number of significant pathways identified in the smaller sample that were consistent
with the full group for LAM (left) and GSA (right).
Middle panels show the number of significant pathways identified in the smaller sample that were inconsistent
with the full group.
Lower panels show the proportion of significant pathways identified that were inconsistent with the full
group.
Left panels correspond to LAM, and right panels to GSA methods.](img/fig4.png)

@@

![**Figure 5. Integration of gene expression data with methylation pathways.**
(A,B) Euler diagrams showing the overlap in statistically significant pathways (FDR<0.05) from gene expression and GSA results (A)
and LAM (B) results.
(C) Multi-dimensional enrichment analysis of promoter and gene body methylation with gene expression using mitch.
Top 20 gene sets shown with the largest S distance after FDR filtering at 0.05.](img/fig5.png)

@@

![**Figure 6. Pathway-level DNA methylation alterations with chronological age.**
(A) Contour heatmap showing the similarity in gene methylation score ranks in the discovery and replication studies.
(B) Mitch pathway enrichment scores (S distances) in discovery and replication studies.
Pathways with MANOVA FDR<0.05 are shown in red while others are shaded grey.
(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
Red indicates increasing methylation and blue shows lower methylation.
(D) An example of a pathway identified with this method, "creation of C4 and C2 activators" shows lower methylation of member genes
in both discovery and replication studies.](img/fig6.png)

@@

![**Figure 7. Pathway-level DNA methylation differences in natural and IVF conceived infants.**
(A) Contour heatmap showing the similarity in gene methylation score ranks in the Estill (HM450K) and Novakovic (EPIC) studies.
(B) Mitch pathway enrichment scores (S distances) in Estil and Novakovic studies.
Pathways with MANOVA FDR<0.05 are shown in red while others are shaded grey.
(C) Heatmap of 30 pathways with most extreme S distances (a surrogate of effect size).
Red indicates higher methylation and blue shows lower methylation.
(D) An example of a pathway identified with this joint enrichment analysis method, "Adrenoreceptors" shows lower methylation of
member genes in both Estill and Novakovic studies.](img/fig7.png)

@@

![**Figure 8. Differential pathway methylation associated with prevalence of 14 common disease states.**
(A) A bar plot showing the number of statistically significant pathways with higher and lower methylation identified in each prevalent condition (FDR<0.05).
(B) A heatmap of S scores for selected pathways across 14 common prevalent disease states.
Stars indicate that the pathway was identified as being among the top five differentially methylated pathways in each direction for each condition.
](img/fig8.png)
